\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-06 paper.tex

% FIXME this is weak, but gives the basic outline
\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG),
although there is some confusion about the interpretation and little
agreement on the concrete details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
The eARG definition is based on the coalescent with recombination
stochastic process, and we show how the assumptions of that process
place limitations on the patterns of inheritance that can be
represented, and that these assumptions are
% Better wording
substantially broken
in modern large-scale datasets.
Representing genetic ancestry
with an eARG also requires complete precision about the ordering
of all recombination events, which we will rarely have sufficient
information to infer.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is disagreement on the
exact object being defined. This causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.
% Thanks Gideon and Alex!
If not addressed, these problems will
impede progress on applying genetic genealogical
information to understand the biological world.

@@

\section*{ARGs and local trees}
Local trees, describing the genetic genealogy of a sample
at a particular location along the genome, are
embedded within in ARG, and the relationship
between the two is subtle and important.
% An ARG is sometimes referred to as being equivalent to a set of
% local trees along the genome [TODO citation], but the connection
% between the two is subtle.
% In order for the mapping to be one-to-one, so that the original
% gARG can be exactly recovered from a set of local trees and
% their genome coordinates,
% some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
which we avoid due the potential confusion with the
statistical concept of marginal distributions
(which could lead to interpretation as an integrated tree over many positions).

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
In all cases the graph visualization is shown on the left
and the corresponding local trees on the right.
Nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent.
(A) A gARG simulated from a backward-time diploid Wright-Fisher
model: genome \noderef{k} has three children, and the genomes
\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple children.
(B) Simplified to remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that never represent coalescences,
i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
(D) Rewrite edges to bypass any unary nodes in local trees.
}
\end{figure}

@@

\section*{Computational efficiency and data interchange}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importantance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of predicted application
% TODO more?
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using,
nowbandegani2023extremely}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems of converting data from one programs
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package,
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.

@@

@@

\section*{Discussion}
Recent breakthroughs have finally made large-scale ARG inference
feasible in practice, leading to a surge of interest
in development of ARG-based inference methods, their evaluation 
and application.
The prospect of ARGs being used routinely within population
and statistical genetics is tantalising,
but in reality there is substantial work to be done to
enable this.
A necessary first step is a degree of terminological clarity.
As discussed in Appendix XXX, the term ``ancestral recombination
graph" has several
subtly different interpretations, depending on context.
The trend to decouple ARGs from their original definiton 
as a stochastic
process and instead use the term as a more general representation of any
recombinant genetic ancestry seems most useful, which we have
tried to clarify and systematise here. Thus
we can think of an ARG as any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which is more limited in what can be described, because of its origins in an idealised mathematical model.

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While it has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
For example, several human datasets now consist of hundreds of thousands of
% TODO cite GeL paper
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and
myriad interconnections of migration, were somehow estimated and
provided as input, ARGs sampled from
the coalescent cannot capture the complexities of realistic family structures. There are many species with sizeable to large number of progeny, making polytomies a necessary feature of local trees. Flexible ARG encoding should accommodate such cases.

Given that there is no single model that can capture the complexities
of modern datasets across species (and taxa; see discussion of SARS-CoV-2
below), we cannot rely on sampling from a distribution to capture
the uncertainty in our ARG estimates. Indeed, even if the coalescent
was a suitable model for datasets such as UKB and we had a means of
sampling from its distribution of ARGs at the scale of hundreds of thousands
of samples, we could only explore the tiniest corner of such an incomprehensibly
large space of possibilities.
Therefore, it is important to systematically
describe and utilise uncertainty about ARG inference.
One approach, enabled by the gARG encoding described here, is to allow
nodes to have (rootward) in-degree greater than two (polytomies
representing uncertainty over the ordering of coalescence events) or 
out-degree greater than two (representing uncertainty over the ordering
of multiple recombination events). Other methods to capture, for example,
uncertainty about node ages and breakpoint locations, is an important
aspect of future work.

@@

% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such precise estimates provide
to downstream applications.
To date, few downstream applications using estimated ARGs
make use of complete recombination information.
Most applications work by examining local trees independently.
For example, the \relate\ selection statistic~\citep{speidel2019method}
computes selection $p$-values by comparing the coalescence rates at various time points.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\citep{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in an ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
of ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
The main application for fully precise ARGs thus far has been
to compute a likelihood under the coalescent with
recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian},
which currently requires the details of all recombination
events to be known.

% Note: haven't said anything about the coalescent here. Do we need to?
The advantages of a model-free representation that naturally
incorporates uncertainty about the ordering of events in and ARG are well illustrated by the
ARGs inferred for SARS-CoV-2 by~\cite{zhan2023towards}.
The COVID-19 pandemic led to collection of whole-genome viral data
at unprecented scale. The GISAID database~\citep{shu2017gisaid}
holds over 15 million sequences as of [FIXME].
Recombination, while rare, is an important
factor in the evolution of
SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing}
with recombinants potentially gaining fitness by combining
mutations from deeply diverged lineages [cite].
\citet{zhan2023towards} inferred two large ARGs for SARS-CoV-2 based
on GISAID data, one containing 1.27 million sequences sampled up to
June 30, 2021 and the other (more sparsely sampled) consisting of
657K sequences sampled up to June 30, 2022.
Sequencing effort at the height of the pandemic produced
tens of thousands of sequences collected in a single day,
with hundreds being identical.
Polytomies are therefore and unescapable reality, and a
fully-resolved binary tree would clearly be false precision.
Because of longitudinal data collection, there is no
requirement that samples must be leaf nodes, and samples
will frequently descend from other (older) samples.
Unary nodes (a node in a local tree with exactly one child)
are also a common feature, when a sample is inferred to have
exactly one direct descendant.
Despite the abundance of data and the rarity of recombination,
the parental lineages cannot be precisely identified
in many cases. For example, many are inferred to have three
or more parental lineages (in agreement with the
analysis resulting in the Pango designation of the recombinant
lineage); in most cases, these will indicate two or more
independent recombination events.
Faithfully representing the uncertainty
around these events seems a more useful approach than
the arbitrary resolution which would be required,
if encoded as an eARG.

% An interesting property of the SARS-CoV-2 dataset is that
% it highlights the uncertainty in the genomic position of
% recombination breakpoints. [Some discussion about why
% we can never know where the breakpoint really happened, and
% some follow-up to note that this is even more acute in
% standard popgen datasets where we're thinking about ancient
% recombs].
% This is a limitation of the
% gARG encoding proposed here, which requires that we specify
% exact intervals of inheritance and therefore precise
% recombination breakpoints.
% An approach in which we can systematically reason about the
% uncertainty of inheritance intervals as they propogate
% through an ARG
% would be an interesting avenue for future research.

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may clarify inference goals and improve
methods for evaluation.
In most cases,
the accuracy of ARG inference is evaluated by comparing some truth simulations
to the inferred ARGs by the pairwise comparison of local trees along the genome
using tree distance
metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping}
as originally proposed by \citet{kuhner2015assessing}.
% While this provides valuable insights into the method's performance,
% it is not clear how well these tree distance metrics reflect
% performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
structure is incorporated in an indirect manner, and the tree
metrics assume unlabelled nodes, effectively discarding information
about node sharing between trees.
Tree distance metrics usually have $O(n^2)$ time complexity and therefore cannot be
applied to the very large sample sizes currently of interest.
A recent trend has been to move away from such tree distance-based approaches and to examine more properties of the inferred ARGs.
\citet{brandt2021evaluation} compared the
distributions of time to the most recent common ancestor across several
methods, which captures many important aspects of dated ARGs
but neglects [SOME OTHER THINGS].
\citet{deng2021distribution} derived the distribution of the expected
waiting distances to tree changes, and compared this to the
outputs of several methods.
[ \citet{ignatieva2023distribution} DID SOMETHING ELSE]
% Not sure this is the right place for this sentence
All of these evaluations have been based on
ground truth data from highly idealised simulations
with small sample sizes,
and the effects of the richness of real data
(beyond simplistic error models)
on modern biobank scale or multi-generational pedigree datasets
are almost entirely unknown.
A suite of true ARG evaluation metrics, that take into account more of the
global topology and can be applied to large ARGs, would be a
valuable addition to the field.

[TODO finish up with a positive note about interchange, and make the point
clearly that tskit is an ARG library. Splice in the stuff about haplotype
blocks and ARGON format here somewhere --- basically this is a derived product
of the gARG encoding, so while there are some computation advantages it's not
fundamental to the actual representation of the thing]

Earlier efforts to standardise ARG interchange
did not succeed~\citep{cardona2008extended,mcgill2013graphml}.

In summary, we advocate for decoupling between generative models that give rise to observed data and data formats that can faithfully and realistically encode these data. We have shown that ``genome'' ARG (gARG) encoding enables flexible encoding of today's datasets with realistic levels of precision. Importantly, genome ARG encoding should facilitate easier data interchange between different tools, which will in turn support active research in this area.

@@

% Optimisation problems to do with ARGs and other phylogenetic networks
There has also been substantial interest in formulating and answering
fundamental questions about properties
of the ARG as a discrete graph structure, abstracting away from the
generating process and focussing on the ARG topology without considering
branch lengths.
The first prominent problem was calculating (lower bounds on) the minimum number of
recombinations required to reconstruct a valid genealogy for a given
sample~\citep{myers2003bounds}, and constructing the corresponding
minimal (parsimonious)
ARGs~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
These problems are NP-hard in general~\citep{wang2001perfect}, and progress has
been achieved through studying various constrained special cases of ARGs~\citep[e.g.][]{gusfield2004optimal} and
other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
focus has been on algorithmic and combinatorial results, that largely do not
have direct relevance to the inference problems described above.

@@

The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
depends on the context in which it is used, and can mean both the
stochastic process generating genealogies in the presence of
recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
wilton2015smc,griffiths2016coalescent},
as well as, more commonly, the concrete realisation of ancestry from a
process~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure, and
argued for using the term to mean specific realised ancestral histories.
Unless otherwise stated, we will use the ARG-as-a-data-structure
interpretation for the remainder of this paper.

@@

Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
of the little ARG process, as well as the extra events which occur in the big ARG process.
Moreover, it depicts the rates of common ancestors and recombination events in each
interval of time of the realisations.
In order to evaluate these rates, and hence the sampling probability
(see e.g.\ \cite[Equation (3)]{mahmoudi2022bayesian}),
it is necessary to know the number of lineages and number of extant links
available for recombination in each time interval.
This information cannot be recovered from the gARG encoding depicted in
e.g.\ Figure \ref{fig-ancestry-resolution}.
For example, it is clear that a recombination takes place between nodes \textsf{b} and
\textsf{d} as well as \textsf{e}, but the exact time of the event is ambiguous,
and thus so is the number of lineages during the time interval.
Conventions can be introduced to resolve such ambiguities;
for the likelihood-based inference algorithms for the coalescent with recombination,
the two parent nodes are typically created at the time of the recombination event.
See the appendix of \cite{baumdicker2021efficient} for details of evaluating
coalescent with recombination likelihoods using this convention.
This is also the interpretation depicted in
Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \textsf{b}
in Figure \ref{fig-ancestry-resolution} should correspond to only one lineage,
along which all $m-1$ links are available for recombination.
The lineage then splits into two at the time of nodes \textsf{d} and \textsf{e}.
Nodes \textsf{k}, \textsf{l}, and \textsf{m} in Figure \ref{fig-ancestry-resolution}
demonstrate that the same issue can affect the number of links available for recombination:
without an external convention, the exact time at which the trapped ancestral material on
node \textsf{k} ceases to be available for an effective recombination in the little ARG process.

>>>>

% FIXME this is weak, but gives the basic outline
\begin{abstract}
New developments have made it possible to infer genetic genealogies in
the presence of recombination at scale, enabling many
downstream applications in population and statistical genetics.
The structure representing such recombinant genetic ancestry
is usually referred to as an ancestral recombination graph (ARG),
although there is some confusion about the interpretation and little
agreement on concrete details.
We propose a concrete definition of ARGs
in terms of genomes and their intervals of genetic inheritance (gARG)
and contrast this with the classical event-based definition (eARG).
The eARG definition is based on the stochatic process of the coalescent with recombination,
and we show how the assumptions of that process
place limitations on the patterns of inheritance that can be
represented, which are inadequate for modern large-scale datasets.
The eARG representation also requires complete precision about the ordering
of all recombination events, which we will rarely have sufficient
information to infer.
We show, in contrast, that the gARG definition
can fully capture the richness of modern large-scale datasets,
enables fine-grained
levels of precision about recombination to be represented,
and forms the basis of an efficient computational framework.
\end{abstract}

@@

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is disagreement on the
exact object being defined. This causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer,gower2022demes}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.
% Thanks Gideon and Alex!
If not addressed, these problems will
impede progress on applying genetic genealogical
information to understand the biological world.

@@

\section*{ARGs and local trees}
% TODO mention Yamato here? I'm sure there were some interesting points 
% about local trees in that paper
Local trees, describing the genetic genealogy of a sample
at a particular location along the genome, are
embedded within in ARG, and the relationship
between the two is subtle and important.
% An ARG is sometimes referred to as being equivalent to a set of
% local trees along the genome [TODO citation], but the connection
% between the two is subtle.
% In order for the mapping to be one-to-one, so that the original
% gARG can be exactly recovered from a set of local trees and
% their genome coordinates,
% some specific properties of the ARG and the trees are required.
Note that many authors use the term ``marginal''
trees~\citep[e.g.][]{griffiths1996ancestral,minichiello2006mapping} [more]
which we avoid due the potential confusion with the
statistical concept of marginal distributions
(which could lead to interpretation as an integrated tree over many positions).

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{illustrations/simplification}
\caption{\label{fig-simplification}
Levels of ARG simplification.
In all cases the graph visualization is shown on the left
and the corresponding local trees on the right.
Nodes are coloured by the number of parents and shaded
according to the proportion of their span over which they are coalescent.
(A) A gARG simulated from a backward-time diploid Wright-Fisher
model: genome \noderef{k} has three children, and the genomes
\noderef{e} and \noderef{g} simultaneously have multiple parents and multiple children.
(B) Simplified to remove all
singly-connected graph components (e.g., diamonds such as \noderef{jlnm}).
(C) Remove nodes that never represent coalescences,
i.e.\ are unary everywhere (e.g.\ \noderef{n} and \noderef{r}).
(D) Rewrite edges to bypass nodes where they appear unary in local trees.
}
\end{figure}

@@

\section*{Computational efficiency and data interchange}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of predicted application
% TODO more?
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using,
nowbandegani2023extremely}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems of converting data from one programs
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package,
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.

@@

[TODO discussion of haplotype block based storage
like ARGON. Basically: this is useful for some applications,
but ultimately it's a derived property of the actual ARG
topology that can be recovered efficiently, as demonstrated by 
\cite{nowbandegani2023extremely}]

@@

\section*{Discussion}
Recent breakthroughs have finally made large-scale ARG inference
feasible in practice, leading to a surge of interest
in development of ARG-based inference methods, their evaluation 
and application.
The prospect of ARGs being used routinely within population
and statistical genetics is tantalising,
but in reality there is substantial work to be done to
enable this.
A necessary first step is a degree of terminological clarity.
As discussed in Appendix XXX, the term ``ancestral recombination
graph" has several
subtly different interpretations, depending on context.
The trend to decouple ARGs from their original definition 
within the context of stochastic
processes and instead use the term as a more general representation of any
recombinant genetic ancestry seems useful, and we have
tried to clarify and systematise it here. Thus
we can think of an ARG as any structure that encodes the
reticulate genetic ancestry of a sample of colinear sequences under
the influence of recombination. The ``genome'' ARG (gARG) encoding
made explicit here is one way we can concretely
define such recombinant ancestry, which we have shown is both
flexible and efficient.
The flexibility of the gARG encoding contrasts with the classical
``event'' ARG (eARG), which is more limited in what can be described, because of its origins in an idealised mathematical model.

Fully decoupling the general concept of an ARG from the coalescent
with recombination stochastic process is an important step.
While it has proven to be a useful and
robust
model~\citep{wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
many modern datasets have properties that grossly
violate its assumptions.
For instance, several human datasets now consist of hundreds of thousands of
% TODO cite GeL paper
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and so sample size is substantially \emph{larger} than the
usually assumed $N_e$ values.
Agricultural datasets are an even more extreme departure from coalescent
assumptions, with hundreds of thousands of samples embedded in
multi-generational pedigrees~\citep{hayes20191000,Ros-Freixedes2020}
and effective population sizes of 100 and even
less~\citep{MacLeod2013,Makanjuola2020,Hall2016,Porcnic2016}.
Even if sufficiently complex demographic models~\citep{gower2022demes}
encompassing hundreds of populations, explosive growth rates and
myriad interconnections of migration, were somehow estimated and
provided as input, ARGs sampled from
the coalescent cannot capture the complexities of realistic family structures. There are many species with sizeable to large number of progeny, making polytomies a necessary feature of local trees. Flexible ARG encoding should accommodate such cases.

Given that there is no single model that can capture the complexities
of modern datasets across species (and taxa; see discussion of SARS-CoV-2
below), we cannot rely on sampling from a distribution to capture
the uncertainty in our ARG estimates. Indeed, even if the coalescent
were a suitable model for datasets such as UKB and we had a means of
sampling from its distribution of ARGs at the scale of hundreds of thousands
of samples, we could only explore the tiniest corner of such an incomprehensibly
large space of possibilities.
Therefore, it is important to systematically
describe and utilise uncertainty about ARG inference.
One approach, enabled by the gARG encoding described here, is to allow
nodes to have (rootward) in-degree greater than two (polytomies
representing uncertainty over the ordering of coalescence events) or 
out-degree greater than two (representing uncertainty over the ordering
of multiple recombination events). Development of other methods to capture, for example,
uncertainty about node ages and recombination breakpoint positions, is an important
aspect of future work.

@@

% % Do downstream applications actually make use of such full ARGs?
Besides the inherent limitations that exist on inferring fully
precise ARGs from data,
we should also consider the value that such precise estimates provide
for downstream applications.
To date, few downstream applications using estimated ARGs
make use of complete recombination information.
Most applications work by examining local trees independently.
For example, the \relate\ selection test~\citep{speidel2019method}
computes selection $p$-values by computing clade size probabilities conditional on the timing of coalescence events in a given local tree.
In their method
for estimating dispersal rates and the locations of genetic
ancestors,
\citep{osmond2021estimating} downsample trees along the genome
so that they can be regarded as approximately independent.
The SIA method for detecting selection~\citep{hejase2022deep}
encodes local trees as a set of lineage counts at discrete
time intervals, and uses these as feature for a
type of machine learning algorithm
that takes ``temporal'' correlations into account. Thus,
while SIA does use some of the information about local tree correlation,
clearly much of the detail about recombination
events in an ARG is lost.
The \texttt{tsdate} algorithm (and related naive estimator
of ancestral location) uses much more between-tree
information~\citep{wohns2022unified}, explicitly using the gARG
encoding to reason about ancestral nodes.
The main application for fully precise ARGs thus far has been
to compute a likelihood under the coalescent with
recombination~\citep[e.g.][]{kuhner2000maximum,mahmoudi2022bayesian},
which currently requires the details of all recombination
events to be known.

% Note: haven't said anything about the coalescent here. Do we need to?
The advantages of a model-agnostic representation that naturally
incorporates uncertainty about the ordering of events in an ARG are well illustrated by~\cite{zhan2023towards}, who
inferred ARGs using millions of SARS-CoV-2 sequences from the
GISAID database~\citep{shu2017gisaid}. In contrast to typical human sequencing datasets, the SARS-CoV-2 data is
sampled continuously through time, sometimes with tens of thousands of sequences collected per day, with relatively little genetic diversity to
distinguish them. The reconstructed ARGs thus contain polytomies (representing
both the uncertainty around coalescence times and the presence of true
multiple mergers), non-leaf sample nodes (sequences with descendants
also present in the dataset) and unary nodes (samples that have
one other sampled descendant). Recombination is known
to be an important factor in the evolution of SARS-CoV-2~\citep{vaninsberghe2021recombinant,jackson2021generation,ignatieva2022ongoing},
and the ARGs also contain recombination nodes. 
Because parental sequences
are generally never sampled themselves, and often 
a recombinant strain is the product of multiple recombination events,
uncertainty around this is captured by recording the ancestry
of each part of the recombinant sequence without arbitrarily assigning
times or orderings for these events. This demonstrates the flexibility of the gARG
encoding in capturing recombinant genealogies regardless of the underlying
(possibly unknowable) generative evolutionary process.

This view of ARGs,
decoupled from generative models and
without the hard requirement
of complete precision on all historical events, may clarify inference goals and improve
methods for evaluation.
In most cases,
ARG inference is evaluated by simulating data from a known ground truth ARG, and comparing this to the inferred version via pairwise comparison of local trees along the genome, for example
using tree distance
metrics~\citep[e.g.][]{robinson1981comparison,kendall2016mapping},
as described by \citet{kuhner2015assessing}.
% While this provides valuable insights into the method's performance,
% it is not clear how well these tree distance metrics reflect
% performance on real data.
In comparing tree-by-tree along the genome, the effects of recombination
are only incorporated in an indirect manner through the correlations
between the local trees, rather than through directly taking into account
the persistence of nodes and edges across multiple trees.
Tree distance metrics usually have $O(n^2)$ time complexity and therefore cannot be
applied to the very large sample sizes currently of interest.
A recent trend has been to move away from such tree distance-based approaches and to examine more properties of the inferred ARGs.
\citet{brandt2021evaluation} focussed on the
distributions of pairwise MRCA times; \citet{deng2021distribution} derived an approximation to the distribution of the 
waiting distances between local trees; \citet{ignatieva2023distribution}
derived the distribution of the genomic span of an edge and that of a
clade of samples. In each case, simulation studies demonstrated 
substantial differences between these quantities in simulated and reconstructed ARGs,
that were not captured using tree-by-tree comparisons.
% Not sure this is the right place for this sentence
These evaluations have been based on
ground truth data from highly idealised simulations, some 
with small sample sizes,
and the effects of the richness of real data
(beyond simplistic error models)
on modern biobank scale or multi-generational pedigree datasets
are almost entirely unknown.
A suite of true ARG evaluation metrics, that take into account more of the
global topology and can be applied to large ARGs, would be a
valuable addition to the field.

Interest in ARG inference methods and downstream applications is 
burgeoning, with exciting developments arriving at ever-increasing pace.
Without agreement on basic terminology and some standardisation
on data formats, however, the ARG revolution may falter.
Here, we have advocated
for decoupling the data structures used to represent ARGs
from generative processes, and have shown that the 
gARG encoding enables flexible description of 
contemporary datasets with varying levels of precision and integral
handling of some important forms of temporal uncertainty.
For ARG based methods to achieve mainstream status, we require
a rich supporting software ecosystem. 
Ideally, this would comprise a wide range of different 
inference methods specialised to different organisms, types 
and scales of data and 
% I'm thinking about things like Espalier specifically wanting
% to characterise the recombs, vs Relate mainly wanting to 
% get good local trees
specific inference goals, sharing a common ARG interchange format. 
The output of these diverse inference methods could then be 
processed by numerous downstream applications, 
enabling users to freely compare the outputs of different 
inference methods on their data, and the result of different
analyses on these inferences.
Earlier efforts to standardise ARG interchange shared 
this vision, 
but did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
Today, however, we have the benefit of a pre-existing ARG
interchange format, already widely used and accompanied 
by high-quality, high-performance software (see Section XXX for details). 
Adopting the \texttt{tskit} library 
as a community standard,
utilising its rich suite of operations,
would substantially accelerate progress, and help 
establish ARG methods as mainstream genetics.

@@

% Optimisation problems to do with ARGs and other phylogenetic networks
There has also been substantial interest in formulating and answering
fundamental questions about properties
of the ARG as a discrete graph structure, focussing on the ARG topology without considering
either branch lengths or indeed the generating process.
The first prominent problem was calculating (lower bounds on) the minimum number of
recombinations required to reconstruct a valid genealogy for a given
sample~\citep{myers2003bounds}, and constructing the corresponding
minimal (parsimonious)
ARGs~\citep{song2003parsimonious,song2005efficient,lyngso2005minimum}.
These problems are NP-hard in general~\citep{wang2001perfect}, and progress has
been achieved through studying various constrained special cases of ARGs~\citep[e.g.][]{gusfield2004optimal} and
other more general types of phylogenetic networks~\citep{huson2010phylogenetic}. The
focus has been on algorithmic and combinatorial results that largely do not
have direct relevance to the inference problems described above.

@@

The goal of this historical overview is to illustrate that the meaning of the term ``ARG" now strongly
depends on the context in which it is used, and can mean both the
stochastic process that generates genealogies in the presence of
recombination~\citep[e.g.][]{nordborg2000linkage,birkner2013ancestral,
wilton2015smc,griffiths2016coalescent},
as well as, more commonly, the concrete realisation of ancestry from a
process~\citep[e.g.][]{gusfield2014recombinatorics,mathieson2020ancestry,brandt2021evaluation}.
\cite{minichiello2006mapping} explicitly distinguished the ARG as a
stochastic process from the ARG as a data structure. We have followed them in
using the term in the latter sense, to mean a specific realised ancestral history.

@@

Figure \ref{hudson_vs_bigARG} illustrates the more complex state space
of the little ARG process, as well as the extra events which occur in the big ARG process.
Moreover, it depicts the rates of common ancestors and recombination events in each
interval of time of the realisations.
In order to evaluate these rates, and hence the sampling probability
(see e.g.\ \cite[Equation (3)]{mahmoudi2022bayesian}),
it is necessary to know the number of lineages and number of extant links
available for recombination in each time interval.
This may not be obvious.  example, in the gARG encoding depicted in Figure \ref{fig-ancestry-resolution}
it is clear that a recombination takes place between nodes \textsf{b} and
\textsf{d} as well as \textsf{e}, but the exact time of the event is ambiguous,
and thus so is the number of lineages during the time interval. This information cannot be recovered
from the gARG encoding 
Conventions can be introduced to resolve such ambiguities;
for the likelihood-based inference algorithms for the coalescent with recombination,
the two parent nodes are typically created at the time of the recombination event.
See the appendix of \cite{baumdicker2021efficient} for details of evaluating
coalescent with recombination likelihoods using this convention.
This is also the interpretation depicted in
Figure \ref{hudson_vs_bigARG}, but it means that the two edges above node \textsf{b}
in Figure \ref{fig-ancestry-resolution} should correspond to only one lineage,
along which all $m-1$ links are available for recombination.
The lineage then splits into two at the time of nodes \textsf{d} and \textsf{e}.
Nodes \textsf{k}, \textsf{l}, and \textsf{m} in Figure \ref{fig-ancestry-resolution}
demonstrate that the same issue can affect the number of links available for recombination:
without an external convention, the exact time at which the trapped ancestral material on
node \textsf{k} ceases to be available for an effective recombination in the little ARG process.
